---
title: 'Homework Assignment Chapter 8 Fitz'
author: "Alex Fitz"
output:
  html_document:
    toc: yes
    toc_depth: 3
  pdf_document:
    toc: yes
    toc_depth: '3'
  word_document: default
---

```{r, warning = FALSE, message=FALSE}
if (!require(BiocManager))
  {install.packages("BiocManager")}
  library(BiocManager)

if (!require(BiocManager::install("multtest")))
  {BiocManager::install("multtest")}
  library(multtest)

if (!require(ALL))
  {install.packages("ALL")}
  library(ALL)

if (!require(lmtest))
  {install.packages("lmtest")}
  library(lmtest)
```

### Question 1: Classification trees with the ALL data.

**(a):** Select the genes with different means over the patient groups with an ANOVA p-value smaller than 0.001.

```{r}

```

**(b):** Using rpart(), find a manner to identify the best 2 genes to predict the B-cell state.

```{r}

```


**(c):** Use rpart() again to construct the best overtall classification tree to predict the B-cell state.


```{r}
```

**(d):** How do the trees and misclassification rates differ?

```{r}
```

***

